% What shall we talk about at ESA?
% Noam Ross
% 14-07-17 20:49:15

`r knitr::opts_chunk$set(cache=TRUE, warning=FALSE, fig.cap="")`


[ESA](http://esa.org/am/) is just around the corner, and many of us are
gearing up and trying to figure out a
[schedule](http://eco.confex.com/eco/2014/schedule/index.cgi) to cover
all the talks and people we can pack in. ESA is a big conference and
there's far too much for any one person to see. In the end, everyone
experiences a [different part of the
elephant](http://en.wikipedia.org/wiki/Blind_men_and_an_elephant).
However, I thought it would be interesting to take a look at the big
picture, and examine the ESA program as a *whole* to see what could be
learned from it. This is the first of (maybe) several posts where I use
some basic text-mining tools to explore the content of the ESA program.


```{r setup, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
require(XML)
require(plyr)
require(tm)
require(dplyr)
require(wordcloud)
require(ggplot2)
if(!require(noamtools)) {
  install_github('noamross/noamtools')
  library(noamtools)
  }
```

First, what are the most common terms in the ESA program?

```{r readfiles,  echo=FALSE, results='hide', cache=TRUE}
# Write all the abstracts into a single text file per year
ppath = "eco.confex.com/eco/2013/webprogram"
paper_files = list.files(ppath, recursive=TRUE, pattern="Paper\\d+\\.html")

cat("", file="corpus/esa2013.txt")
a_ply(paper_files, 1, function(paper) {
  paper_xml = htmlTreeParse(file.path(ppath, paper), useInternalNodes = TRUE, trim=TRUE)
  ab = try(xmlValue(paper_xml[['//div[@class="abstract"]']]), silent = TRUE)
  if(class(ab) != "try-error") {
  ab = stringi::stri_replace_all_fixed(ab, "Background/Question/Methods", "")
  ab = stringi::stri_replace_all_fixed(ab, "Results/Conclusions", "")
  cat(ab, file="corpus/esa2013.txt", append=TRUE)
  }
}, .progress = "time")
cat("\n", file="corpus/esa2013.txt", append=TRUE)

ppath = "eco.confex.com/eco/2014/webprogram"
paper_files = list.files(ppath, recursive=TRUE, pattern="Paper\\d+\\.html")

cat("", file="corpus/esa2014.txt")
a_ply(paper_files, 1, function(paper) {
  paper_xml = htmlTreeParse(file.path(ppath, paper), useInternalNodes = TRUE, trim=TRUE)
  ab = try(xmlValue(paper_xml[['//div[@class="abstract"]']]), silent = TRUE)
  if(class(ab) != "try-error") {
  ab = stringi::stri_replace_all_fixed(ab, "Background/Question/Methods", "")
  ab = stringi::stri_replace_all_fixed(ab, "Results/Conclusions", "")
  cat(ab, file="corpus/esa2014.txt", append=TRUE)
  }
}, .progress = "time")
cat("\n", file="corpus/esa2014.txt", append=TRUE)
```


```{r loadcorpus, echo=FALSE, results='hide'}
#Load and clean the corpus
esa = Corpus(DirSource("corpus/")) %>%
  tm_map(stripWhitespace) %>%
  tm_map(tolower) %>%
  tm_map(removeWords, stopwords("english")) %>%
  tm_map(removeNumbers) %>%
  tm_map(removePunctuation, preserve_intra_word_dashes=TRUE)

esa2 = esa %>% tm_map(stemDocument)

tfa = TermDocumentMatrix(esa, list(bounds=list(local=c(8, Inf))))

dict = Terms(tfa)[!(Terms(tfa) %in% c("special", "specialist", "specialists", "ecologic"))]
esaComp = TermDocumentMatrix(esa2, list(bounds=list(local=c(20, Inf))))
newTerms = stemCompletion(Terms(esaComp), dict, 'shortest')
esaComp$dimnames$Terms = newTerms
```

```{r plot1, echo=FALSE, fig.width=5, fig.height=8, cache=FALSE}
freq <- sort(slam::row_sums(esaComp), decreasing=TRUE)
words <- names(freq) 
wd = data.frame(words=factor(words[1:30], levels=words[30:1]), freq=freq[1:30])
ggplot(wd, aes(x=words, y=freq, label=words)) +
  geom_bar(stat='identity') +
  geom_text(hjust = 1.05, col='white') +
  coord_flip() +
  theme_nr +
  theme(
       axis.title.y=element_blank(),
        axis.ticks.y=element_blank(),
       axis.text.y=element_blank(),
        axis.title.x=element_text(size=18),
        panel.grid=element_blank()) +
  ylab('Number of Uses in ESA Program') + 
  ggtitle('Top 30 Terms in ESA 2014 Abstracts')
```

Few surprises here. "Species" would have been my guess for the top.
"Plants" are probably on top because ecologists usually refer to animals
by various sub-groups. The rest are fairly ho-hum: ecology and science-y
words.

It's more interesting to ask how the topics at ESA *change*
from year to year. Below I show the terms whose use in ESA abstracts
changed the most between 2013 to 2014:

```{r mung, echo=FALSE, fig.width=8, fig.height=11, cache=FALSE} 
dat = as.data.frame(as.matrix(esaComp))
colnames(dat) = c("esa13", "esa14")
dat$term = rownames(dat)
dat = dat %>% mutate(p13 = esa13/sum(esa13), p14=esa14/sum(esa14)) %>%
  mutate(chg = (p14-p13)) %>%
  mutate(pctchg = (p14-p13)/p13) %>% 
  mutate(dir = factor(ifelse(chg > 0, 2014, 2013))) %>%     
  arrange(-abs(chg))

dat2 = head(dat, 50) %>% arrange(chg) %>% droplevels
dat2$term = factor(dat2$term, levels=dat2$term[order(dat2$chg)])
ggplot(dat2, aes(x=term, label=term, y=chg*1000, fill=dir, hjust=ifelse(dir==2014, -0.05, 1.05))) +
  geom_bar(stat="identity", col="white", lwd=0.5, alpha=0.5) +
  geom_text(col="black", y=0, size=5) +
  coord_flip() +
  scale_fill_manual(values=c("red", "darkgreen"), guide=FALSE) +
#  scale_color_manual(values=c("red", "green"), guide=FALSE) +
  theme_nr +
  theme(panel.grid.major.y=element_blank(),
        axis.title.x=element_text(size=18),
        axis.title.y=element_blank(),
        axis.text.y=element_blank()
 #       panel.grid.major.x=element_line(colour="grey"),
  #      panel.grid.minor.x=element_line(colour="grey")
        ) +
  ggtitle('50 terms with  greatest change in the ESA Program, 2013-2014') +
  ylab('Change in frequency (occurrence per 1000 words)')
```

This paints a much more interesting picture. The rise of California and
the fall of Minnesota make sense given the change in the meeting's
location. But we can see the influence of landscape on topics as well.
We see fewer words associated with freshwater ecosystems, prairies, and
forests this year, and more associated with fire and other plant
systems. Also, we see a difference in the *kinds* of ecology in the
program. This year there are fewer words like "biomass" and "nutrient" -
those common in ecosystem ecology - and more like "pollinator"[^1],
"phenology", and "network" - those associated with the study of species
interactions.

It's possible that these changes are due to changes in what's popular in
ecology, but it is also likely that many of the concepts captured in
these terms - ecosystem, community, and landscape ecology - are
influenced by region. After all, an ecosystem perspective is likely to
dominate in the Midwest, where an abundance of lakes have been important
in the research of freshwater nutrient cycling, and a landscape
perspective may be important in California, which has such heterogeneity
of habitats. This is a pretty good argument for keeping ESA's location
moving, so that no regional perspective dominates every year.

These are the biggest *changes*, but have the biggest *topics* changed? The
plot below is similar to that above, but instead of plotting the words with the
greatest absolute change, I plot the change of the 50 words that are most
common across both years:

```{r big, echo=FALSE, fig.width=8, fig.height=11, cache=FALSE}

dat = dat %>% mutate(rank = rank(-esa13 - esa14), r13 = rank(-esa13), r14=rank(-esa14))

dat3 = dat %>% filter(rank <= 50) %>% droplevels %>% arrange(term)
dat3$term = factor(dat3$term, levels=dat3$term[order(dat3$chg)])
ggplot(dat3, aes(x=term, label=term, y=chg*1000, fill=dir, hjust=ifelse(dir==2014, -0.05, 1.05))) +
  geom_bar(stat="identity", col="white", lwd=0.5, alpha=0.5) +
  geom_text(col="black", y=0, size=5) +
  coord_flip() +
  scale_fill_manual(values=c("red", "darkgreen"), guide=FALSE) +
#  scale_color_manual(values=c("red", "green"), guide=FALSE) +
  theme_nr +
  theme(panel.grid.major.y=element_blank(),
        axis.title.x=element_text(size=18),
        axis.title.y=element_blank(),
        axis.text.y=element_blank()
        ) +
  ggtitle('Change in the 50 most common terms in the ESA Program, 2013-2014') +
  ylab('Change in frequency (occurrence per 1000 words)')
```


This is somewhat less clear. If one squints, one could argue that there
are more words associated with species interactions, environmental
change, and management at the top, and more words associated with
forests at the bottom.  Words in the middle ("ecology", "community") are consistently popular across both years.  Finally, perhaps significance is falling out of fashion?

That's just a quick first pass. I haven't yet thought much about how one models these data to understand effect sizes and significance. I welcome suggestions for further
analyses and better ways to plot/organize this data. Check out [this repository on
github](https://github.com/noamross/esaprog) for the code that generated
these plots and how to grab the ESA program text for your own use. See
you in a few weeks!

**P.S.** While messing with the ESA program text, I also created [\@esa_titles](http://twitter.com/esa_titles), a twitter account that re-mixes
ESA talk titles. Have a look for talks you wish you could see. :)

[^1]: "pollinia" is a stand-in for all pollination-related words here,
    as I applied [stemming](http://en.wikipedia.org/wiki/Stemming) to
    the text.